---
title: "Report on Cyclist Trips 2019"
author: "Alexander Bryl"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install}
options(scipen=999) 
install.packages("tidyverse") 
library(tidyverse)
#helps wrangle data \# Use the conflicted package to manage conflicts
library(conflicted)

# Set dplyr::filter and dplyr::lag as the default choices

conflict_prefer("filter", "dplyr") conflict_prefer("lag", "dplyr")
```

# STEP 1: COLLECT DATA

### Upload Divvy datasets (csv files) here

```{r read data}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv") 
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
```

## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

### Compare column

names each of the files. While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r look}
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```

### Rename columns to make them consistent with q1_2020 (as this will be the supposed going-forward table design for Divvy)

```{r rename}
(q1_2019 <- rename(q1_2019 ,ride_id = trip_id   ,rideable_type = bikeid
,started_at = start_time ,ended_at = end_time ,start_station_name =
from_station_name ,start_station_id = from_station_id ,end_station_name
= to_station_name ,end_station_id = to_station_id ,member_casual =
usertype ))

(q2_2019 <- rename(q2_2019 ,ride_id = "01 - Rental Details Rental ID"
,rideable_type = "01 - Rental Details Bike ID"
,started_at = "01 - Rental Details Local Start Time"
,ended_at = "01 - Rental Details Local End Time" ,start_station_name =
"03 - Rental Start Station Name" ,start_station_id = "03 - Rental Start
Station ID" ,end_station_name = "02 - Rental End Station Name"
,end_station_id = "02 - Rental End Station ID"
,member_casual = "User Type" ,gender = "Member Gender" ,birthyear =
"05 - Member Details Member Birthday Year" ,tripduration = "01 - Rental
Details Duration In Seconds Uncapped" ))

(q3_2019 <- rename(q3_2019 ,ride_id = trip_id   ,rideable_type = bikeid
,started_at = start_time ,ended_at = end_time ,start_station_name =
from_station_name ,start_station_id = from_station_id ,end_station_name
= to_station_name ,end_station_id = to_station_id ,member_casual =
usertype ))

(q4_2019 <- rename(q4_2019 ,ride_id = trip_id ,rideable_type = bikeid
,started_at = start_time ,ended_at = end_time ,start_station_name =
from_station_name ,start_station_id = from_station_id ,end_station_name
= to_station_name ,end_station_id = to_station_id ,member_casual =
usertype ))

```

### Convert ride_id and rideable_type to character so that they can stack correctly

```{r convert}
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id)
,rideable_type = as.character(rideable_type))

q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id)
,rideable_type = as.character(rideable_type))

q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id)
,rideable_type = as.character(rideable_type))

q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id)
,rideable_type = as.character(rideable_type))
```

### Stack individual quarter's data frames into one big data frame

```{r stack}
all_trips <- bind_rows(q1_2019, q2_2019, q3_2019, q4_2019)
```

### Remove birthyear, and gender fields as this data was dropped beginning in 2020

```{r remove}
all_trips <- all_trips %\>% 
select(-c(birthyear, gender))
```

## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

#### There are a few problems we will need to fix:

-   

    (1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

-   

    (2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

-   

    (3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

-   

    (4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

#### Reassign to the desired values (we will go with the current 2020 labels)

```{r reassign}
all_trips <- all_trips %>% mutate(member_casual = recode(member_casual
,"Subscriber" = "member" ,"Customer" = "casual"))

```

#### Add columns that list the date, month, day, and year of each ride.

This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

```{r add_columns }
all_trips$date <- as.Date(all_trips$started_at) all_trips$month_num <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#### Add a "ride_length" calculation to all_trips (in seconds)

```{r ride_len}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```

#### Remove "bad" data

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

```{r clean}
all_trips_v2 <-
all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

```{r summary}
summary(all_trips_v2$ride_length)
```

#### Compare members and casual users

```{r compare}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN =
mean) aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual,
FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN =
max) aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual,
FUN = min)
```

#### Let's visualize the number of rides by rider type

```{r viz_num_rides}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill =
member_casual)) + 
  geom_col(position = "dodge") + 
  labs(title = "Week Days VS. Number of Rides", 
  subtitle = "In 2019") +
  labs(x = "Day of the Week", y = "Number of Rides")
```

#### Let's create a visualization for average duration

```{r avg_duration}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
  ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday,
             y = average_duration,
             fill= member_casual)) + 
  geom_col(position = "dodge") + 
  labs(title = "Average trip Duration vs Day of the Week", 
  subtitle = "In 2019") +
  labs(x = "Day of the Week",
       y = "Average Duration (min.)")
```
